Three distinct patterns of mental response after a sport accident

### 2023-02-27 paper/markdown/manuscript.Rmd
Feedback changes
Changes in text as proposed by Katharina

author: 
  - Katharina Hüfner, MD:
      institute: [psych]
      email: Katharina.Huefner@i-med.ac.at
      correspondence: true
  - Hanna Salvotti, MD:
      institute: [psych]
  - Piotr Tymoszuk, PhD:
      institute: [daas]

@@

keywords: "sport accident, mental health, post-traumatic syndrome disorder, post-traumatic growth"

@@

__Methods:__ A cross-sectional survey study was conducted with 307 individuals treated at the Innsbruck University Hospital (Austria) following a non-professional sport accident. 
Measures of anxiety (GAD-7), depression, somatic symptoms and panic (PHQ), resilient comping (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS), post-traumatic growth (PTGI) and post-traumatic syndrome (PTSD) were recorded along with demographic, socioeconomic, medical history, accident- and injury-related data. 
Patterns of mental readouts were defined by semi-supervised clustering and factors affecting the cluster assignment were identified by statistical testing and conditional random forest machine learning.

__Results:__ Participants were assigned to three mental clusters characterized by a neutral mental response, post-traumatic growth and symptoms of post-traumatic syndrome disorder, respectively. 
Overall differences in demographic, socioeconomic, injury- and accident-related variables between the clusters were low-to-moderate. 
The post-traumatic growth and post-traumatic syndrome clusters were characterized by relatively frequent hospitalization, surgery, somatic illness and female gender. 
The most predictive factors for the post-traumatic syndrome cluster were more cautious behavior and flashbacks during sport activity, low income, mental disorders, somatic consequences of the accident and subjective need for psychological support.

__Conclusion:__ A subset of non-professional sport accident victims develops signs of mental health problems including symptoms of post-traumatic syndrome disorder. 
Patients with low socioeconomic status, somatic and mental conditions, and hospitalization represent a vulnerable group which may benefit most from a psychological intervention.

@@

sport accident, mental health, post-traumatic syndrome disorder, post-traumatic growth

@@

The psychometric battery (__Supplementary Table S2__) consisted of German versions of assessment tools for 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
resilient coping (RS13: 13-item resilience scale) [@Leppert2008], 
loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
loss of quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
post-traumatic syndrome disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
The tools displayed good-to-excellent consistency as measured by McDonald's $\omega$ > 0.8 [@McDonald1999] (__Supplementary Table S3__).

Symptoms of anxiety were defined as GAD-7 $geq$ 10, symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], significant somatization/persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
Resilient coping classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008].

@@

The training subset was clustered in respect to normalized median-centered numeric psychometric scores by the partition around medoids algorithm with the cosine distance measure between observations [@Drost2018; @Schubert2019]. 
The clustering procedure was characterized a good explanatory performance (explained clustering variance: ratio of between-cluster sum of squares to total sum of squares) and the superior stability in 10-fold cross-validation [@Lange2004] as compared with hierarchical and KMEANS algorithms. 
The cluster number was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic [@Kassambara2020]. 
Three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder) cluster. 
Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
The quality of cluster predictions was assessed by comparing explained clustering variances in the training and test subset, visual inspection of UMAP (uniform manifold approximation and projection) layouts [@McInnes2018; @Konopka2022], of distance heat maps and heat maps of normalized clustering feature levels.

@@

As compared with the study participants, the individuals excluded from the analysis had lower employment rates, household income and less chronic somatic conditions. 
The excluded collective tended towards more biking and less ski/snowboard accidents and less professional rescue need. 
Their injuries were on average less severe as measured by AIS [@Gennarelli2006] with more often upper limb injuries, and lower hospitalization and surgery rates. 
The excluded individuals tended suffered less often from somatic health consequences of the accident and tended towards lower ratings of post-traumatic syndrome disorder but also diminished post-traumatic growth and less sense of coherence. 
These differences were significant or near-significant with weak effects sizes (V $\leq$ 0.22 or r $\leq$ 0.15). 
This analysis suffered also from high data missingness in the excluded collective (__Supplementary Table S4__).

In the analyzed cohort, the accidents happened between 
`r ptsd$dataset$accident_date %>% range(na.rm = TRUE) %>% paste(collapse = ' and ')` and were managed at the study site between `r ptsd$dataset$ward_date %>% range(na.rm = TRUE) %>% paste(collapse = ' and ')`. 
The median time between the ward management and survey completion was 
`r median(ptsd$dataset$obs_time)` days (interquartile range: 
`r quantile(ptsd$dataset$obs_time, probs = c(0.25, 0.75)) %>% paste(collapse = ' - ')`). 
The participants were predominantly middle-aged (median `r median(ptsd$dataset$age)` years) and 
`r signif(table(ptsd$dataset$sex)/sum(table(ptsd$dataset$sex)), 2)[1] * 100`% of them were females. 
The vast majority had secondary or tertiary education grade and two-third were professionally active. 
Sport and trauma-risk professions constituted only a minor fraction of the cohort (< 8%). 
Over 40% of participants had high annual household incomes of $\geq$ 45000 Euro. 
Less than 10% of participants were smokers or alcohol/drug consumers. 
Frequencies of somatic and mental conditions were 
`r signif(table(ptsd$dataset$somatic_comorbidity)/sum(table(ptsd$dataset$somatic_comorbidity)), 2)[2] * 100`% and 
`r signif(table(ptsd$dataset$psych_comorbidity)/sum(table(ptsd$dataset$psych_comorbidity)), 2)[2] * 100`%, respectively. 
The rates of smoking (Austria: 20.6%) and somatic (38.3%) illness were lower than in the general population [@StatistikAustria2022a; @StatistikAustria2022]. 
Four of ten participants were directly affected by or witnessed a traumatic event prior to the accident as gauged by the DIA-X questionnaire [@Maercker2005].

@@

Clinically relevant signs of 
anxiety (`r signif(table(ptsd$dataset$gad7_total_class)/sum(table(ptsd$dataset$gad7_total_class)), 2)[2] * 100`%), 
depression (`r signif(table(ptsd$dataset$phq9_total_class)/sum(table(ptsd$dataset$phq9_total_class)), 2)[2] * 100`%) and 
somatization (`r signif(table(ptsd$dataset$phq_events_total_class)/sum(table(ptsd$dataset$phq_events_total_class)), 2)[2] * 100`%) were rare in the study cohort. 
The median loss of quality of life was scored with `r median(ptsd$dataset$eurohis_total) %>% signif(2)` EUROHIS QOL points, which was substantially more than in a generalized Western European population [@Schmidt2006]. 
The participants were most satisfied with their overall quality of life and housing (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1B__). 
Roughly two-third of participants were assigned to the high resilient coping class [@Leppert2008]. 
The scores of overall post-traumatic growth (median: `r median(ptsd$dataset$ptgi_total) %>% signif(2)`) and its specific domains were lower in the study cohort as compared with the figures originally reported for the PTGI tool [@Tedeschi1996] (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1C__). 
In `r signif(table(ptsd$dataset$dsm5_cluster_class)/sum(table(ptsd$dataset$dsm5_cluster_class)), 2)[2] * 100`% of participants symptoms of post-traumatic syndrome disorder defined as positivity for at least one PCL-5 DSM-5 tool domain [@Bovin2016] could be observed. 
Among them, domain B symptoms (`r signif(table(ptsd$dataset$dsm5_B_class)/sum(table(ptsd$dataset$dsm5_B_class)), 2)[2] * 100`%) were the most frequent and domain D symptoms were the rarest (`r signif(table(ptsd$dataset$dsm5_D_class)/sum(table(ptsd$dataset$dsm5_D_class)), 2)[2] * 100`%) (__Table \@ref(tab:tab-cohort-mental)__ and __Supplementary Figure S1D__).

@@

To comprehend types of mental response to the sport accident, we subjected the study participants to semi-supervised clustering in respect to measures of anxiety, depression, panic, persistent somatic symptoms, resilience, sense of coherence, quality of life, post-traumatic growth and post-traumatic syndrome disorder. 
Among several clustering algorithms compared in the training subset, partition around medoids with cosine distance between observations [@Drost2018; @Schuber2019] demonstrated good explanatory power and superior reproducibility in cross-validation [@Lange2004] (__Supplementary Figure S2A__). 
Three clusters of participants named neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder) cluster after the patterns of their characteristic mental features were identified in the training subset. 
Subsequently, the mental cluster assignment could be robustly validated in the test subset as evident from comparable fractions of explained clustering variance in both study cohort subsets (training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`) and comparable cluster sizes (__Supplementary Figures S1B__ - __S4__).

Roughly one-third study participants assigned to the neutral clusters were characterized by low scores of anxiety, depression, panic and somatic symptoms along with high rating of coherence, resilient coping and quality of life. 
Scores of post-traumatic growth and post-traumatic syndrome were the lowest in the neutral cluster. 
The PTG cluster demonstrated similarly low score of major mental health problems as neutral cluster individuals. 
By contrast, the overall measures of post-traumatic growth and its domains were the highest in the PTG cluster. 
In addition, ratings of post-traumatic syndrome disorder symptoms were slightly elevated in the PTG as compared with the neutral cluster. 
The remaining participants subsumed in the PTSD cluster displayed the most intense signs of anxiety, depression, panic and somatic symptoms as well as poor sense of coherence, resilient coping and low quality of life as compared with the neutral and PTG cluster. 
PTSD cluster was hallmarked by the highest scores of post-traumatic syndrome disorder symptoms in every PCL-5 DSM-5 domain. 
Post-traumatic growth was also present in the PTSD cluster as evident from elevated PTGI scores (__Figure \@ref(fig:fig-clust-psych)__, __Supplementary Figure S4__, __Supplementary Table S6__).

Along this line, clinically relevant symptoms of anxiety, depression and somatization were present virtually only in the PTSD clusters of the training and test subsets of the study cohort. 
Furthermore, frequencies of low and moderate resilient coping classes were by far the highest in the PTSD cluster. 
Finally, approximately 40% of PTSD cluster individuals were screened positive for at least one post-traumatic syndrome disorder symptom which was more than in the PTG 
(training: `r signif(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)), 2)[2] * 100`%, 
test: `r signif(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)), 2)[2] * 100`%) or neutral cluster 
(training: `r signif(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)), 2)[2] * 100`%, 
test: `r signif(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)), 2)[2] * 100`%). 
An analogical trend was observed for the remaining post-traumatic syndrome disorder domains (__Figure \@ref(fig:fig-clust-symptoms)__).

@@

In individuals surveyed on average 3.5 years after the non-professional sport accident, some of its consequences like somatic health problems, unwilling flashbacks as well as phenomena of post-traumatic growth and symptoms of post-traumatic syndrome disorder were still tangible. 
By semi-supervised clustering we could discern three distinct patterns of mental health in this collective with each present in approximately one-third of participants: (1) a neutral mental response reaction, (2) salutatory reaction to the accident characterized by post-traumatic growth and (3) mental health deterioration hallmarked by symptoms of anxiety, depression, somatization, resilience loss and signs of post-traumatic syndrome disorder. 
Notably, separation of the PTG and PTSD clusters was in part blurry. 
Symptoms of post-traumatic syndrome were observed also in PTG cluster participants albeit less frequently than in the PTSD cluster. 
Conversely, mental health problems in PTSD cluster participants were paralleled by some signs of post-traumatic growth. 
In a detailed comparison of `r length(clust_bcg$variables)` factors between the clusters, no distinct demographic, socioeconomic, clinical or accident-related background of the mental clusters could be identified. 
We found a moderate enrichment of females, low-to-moderate income participants, non-academic graduates, chronic somatic and psychiatric conditions as well as individuals requiring hospitalization and surgery in the PTG and, in particular, in the PTSD cluster. 
Injury severity and localization or professional rescue rates indicative of potentially more severe accident were comparable between the clusters. 
Similarly, the cluster assignment was unlikely affected by psychological support after the accident. 
In turn, the PTSD and, to a lesser extent, the PTG cluster were characterized by somatic health consequences of the accident, flashbacks, more caution during sport activity and subjective need for psychological support. 
Finally, the lacking non-mental explanatory variables strongly associated with the cluster assignment likely resulted in sub-optimal performance of machine learning classifiers employed for prediction of the cluster assignment and identification of accident victims at risk of mental health problems.

@@

Our study bears limitations. 
First, although the key phenomena phenomena in the newly identified mental clusters, post-traumatic growth and post-traumatic syndrome disorder, are clearly associated with trauma [@Tedeschi1996; @Bovin2016], we could not exclude the mental cluster assignment to be affected by other events than the sport accident. 
To account for that, follow-up studies including a control group are needed. 
Second, our variable set misses potentially important explanatory factors for the mental cluster classification such as length of hospital stay, rehabilitation need or ability to work. 
Finally, effects of the accident, injury severity, hospitalization or surgery on mental health after an accident may have been obscured in the cross-sectional cohort and needs to be investigated in a more defined collective e.g. of high risk sport accident victims or hospitalized individuals.

@@

tables$cohort_mental %>% 
  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
  flextable %>% 
  flextable::width(1, width = 5.5, unit = 'cm') %>% 
  flextable::width(2, width = 7.5, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations. Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier. Three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic syndrome disorder), All clustering parameters differed significantly between the clusters as assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic. Mean normalized scores of major mental health readouts (A), diminished quality of life (B), post-traumatic growth (C) and post-traumatic symptom disorder (D) in the mental clusters of the training and test subsets of the study cohort are depicted as solid lines. Tinted regions represent 2 $\times$ SEM (standard error of the mean). Numbers of observations in the clusters are displayed in the plot captions. GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic syndrome disorder._

@@

_Distribution of gender (A), annual household income (B), and frequencies of self-reported somatic (C) and mental illness (D) in the mental clusters. Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic. Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots. Effect sizes and p-values are displayed in the plot captions. Numbers of observations in the clusters are presented in the X axes._

>>>>

author: 
  - Katharina Hüfner, MD:
      institute: [psych]
      email: Katharina.Huefner@tirol-kliniken.at
      correspondence: true
  - Hanna Salvotti, MD:
      institute: [psych]
  - Piotr Tymoszuk, PhD:
      institute: [daas]

@@

keywords: "sport accident, mental health, post-traumatic stress disorder, post-traumatic growth"

@@

__Methods:__ A cross-sectional survey study was conducted with `r nrow(ptsd$dataset)` individuals treated at the Innsbruck University Hospital (Austria) following a non-professional sport accident. 
Measures of anxiety (GAD-7), depression, somatic symptoms and panic (PHQ), resilient comping (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS), post-traumatic growth (PTGI) and post-traumatic stress disorder (PTSD) were recorded along with demographic, socioeconomic, medical history, accident- and injury-related data. 
Patterns of mental readouts were defined by semi-supervised clustering and factors affecting the cluster assignment were identified by conditional random forest machine learning.

__Results:__ Participants were assigned to three mental clusters characterized by a neutral mental response, post-traumatic growth and post-traumatic stress symptoms, respectively. 
Overall differences in demographic, socioeconomic, injury- and accident-related variables between the clusters were weak-to-moderate. 
The post-traumatic growth and post-traumatic stress clusters were characterized by relatively frequent hospitalization, surgery, pre-existing somatic illness and female gender. 
The key characteristics the post-traumatic stress cluster were more caution and flashbacks during sport activity, low income, pre-existing mental disorders, somatic consequences of the accident and subjective need for psychological support.

__Conclusion:__ A subset of non-professional sport accident victims develops signs of mental health problems including symptoms of post-traumatic stress disorder. 
Patients with low socioeconomic status, pre-existing somatic and mental conditions, and hospitalization represent a vulnerable group which may benefit most from a psychological intervention.

@@

sport accident, mental health, post-traumatic stress disorder, post-traumatic growth

@@

The psychometric battery (__Supplementary Table S2__) consisted of German versions of assessment tools for 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
resilient coping (RS13: 13-item resilience scale) [@Leppert2008], 
loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
loss of quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
The tools displayed good-to-excellent consistency as measured by McDonald's $\omega$ > 0.8 [@McDonald1999] (__Supplementary Table S3__).

Symptoms of anxiety were defined as GAD-7 $\geq$ 10, symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], significant somatization/persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
Resilient coping classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008].

@@

The training subset was clustered in respect to normalized median-centered numeric psychometric scores by the partition around medoids algorithm with the cosine distance measure between observations [@Drost2018; @Schubert2019]. 
The clustering procedure was characterized a good explanatory performance (explained clustering variance: ratio of between-cluster sum of squares to total sum of squares) and the superior stability in 10-fold cross-validation [@Lange2004] as compared with hierarchical and KMEANS algorithms. 
The cluster number was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic [@Kassambara2020]. 
Three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder) cluster. 
Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 7-nearest neighbor classifier. 
The quality of cluster predictions was assessed by comparing explained clustering variances in the training and test subset, visual inspection of UMAP (uniform manifold approximation and projection) layouts [@McInnes2018; @Konopka2022], of distance heat maps and heat maps of normalized clustering feature levels.

@@

As compared with the study participants, the individuals excluded from the analysis had lower employment rates, household income and less chronic somatic conditions. 
The excluded collective tended towards more biking and less ski/snowboard accidents and less professional rescue need. 
Their injuries were on average less severe as measured by AIS [@Gennarelli2006] with more often upper limb injuries, and lower hospitalization and surgery rates. 
The excluded individuals tended suffered less often from somatic health consequences of the accident and tended towards lower ratings of post-traumatic stress disorder but also diminished post-traumatic growth and less sense of coherence. 
These differences were significant or near-significant with weak effects sizes (V $\leq$ 0.22 or r $\leq$ 0.15). 
This analysis suffered also from high data missingness in the excluded collective (__Supplementary Table S4__).

In the analyzed cohort, the accidents happened between 
`r ptsd$dataset$accident_date %>% range(na.rm = TRUE) %>% paste(collapse = ' and ')` and were managed at the study site between `r ptsd$dataset$ward_date %>% range(na.rm = TRUE) %>% paste(collapse = ' and ')`. 
The median time between the emergency unit or ward management and survey completion was 
`r median(ptsd$dataset$obs_time)` days (interquartile range: 
`r quantile(ptsd$dataset$obs_time, probs = c(0.25, 0.75)) %>% paste(collapse = ' - ')`). 
The participants were predominantly middle-aged (median `r median(ptsd$dataset$age)` years) and 
`r signif(table(ptsd$dataset$sex)/sum(table(ptsd$dataset$sex)), 2)[1] * 100`% of them were females. 
The vast majority had secondary or tertiary education grade and two-third were professionally active. 
Sport and trauma-risk professions constituted only a minor fraction of the cohort (< 8%). 
Over 40% of participants had high annual household incomes of $\geq$ 45000 Euro. 
Less than 10% of participants were smokers or alcohol/drug consumers. 
Frequencies of somatic and mental conditions were 
`r signif(table(ptsd$dataset$somatic_comorbidity)/sum(table(ptsd$dataset$somatic_comorbidity)), 2)[2] * 100`% and 
`r signif(table(ptsd$dataset$psych_comorbidity)/sum(table(ptsd$dataset$psych_comorbidity)), 2)[2] * 100`%, respectively. 
The rates of smoking (Austria: 20.6%) and somatic (38.3%) illness were lower than in the general population [@StatistikAustria2022a; @StatistikAustria2022]. 
Four of ten participants were directly affected by or witnessed a traumatic event prior to the accident as gauged by the DIA-X questionnaire [@Maercker2005].

@@

Clinically relevant signs of 
anxiety (`r signif(table(ptsd$dataset$gad7_total_class)/sum(table(ptsd$dataset$gad7_total_class)), 2)[2] * 100`%), 
depression (`r signif(table(ptsd$dataset$phq9_total_class)/sum(table(ptsd$dataset$phq9_total_class)), 2)[2] * 100`%) and 
somatization (`r signif(table(ptsd$dataset$phq_events_total_class)/sum(table(ptsd$dataset$phq_events_total_class)), 2)[2] * 100`%) were rare in the study cohort. 
The median loss of quality of life was scored with `r median(ptsd$dataset$eurohis_total) %>% signif(2)` EUROHIS QOL points. 
The overall quality of life was hence substantially worse than in a generalized Western European population [@Schmidt2006]. 
The participants were most satisfied with their overall quality of life and housing (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1B__). 
Roughly two-third of participants were assigned to the high resilient coping class [@Leppert2008]. 
The scores of overall post-traumatic growth (median: `r median(ptsd$dataset$ptgi_total) %>% signif(2)`) and its specific domains were lower in the study cohort as compared with the figures originally reported for the PTGI tool [@Tedeschi1996] (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1C__). 
In `r signif(table(ptsd$dataset$dsm5_cluster_class)/sum(table(ptsd$dataset$dsm5_cluster_class)), 2)[2] * 100`% of participants symptoms of post-traumatic stress disorder defined as positivity for at least one PCL-5 DSM-5 tool domain [@Bovin2016] could be observed. 
Among them, domain B symptoms (`r signif(table(ptsd$dataset$dsm5_B_class)/sum(table(ptsd$dataset$dsm5_B_class)), 2)[2] * 100`%) were the most frequent and domain D symptoms were the rarest (`r signif(table(ptsd$dataset$dsm5_D_class)/sum(table(ptsd$dataset$dsm5_D_class)), 2)[2] * 100`%) (__Table \@ref(tab:tab-cohort-mental)__ and __Supplementary Figure S1D__).

@@

To comprehend types of mental response to the sport accident, we subjected the study participants to semi-supervised clustering in respect to measures of anxiety, depression, panic, persistent somatic symptoms, resilience, sense of coherence, quality of life, post-traumatic growth and post-traumatic stress disorder. 
Among several clustering algorithms compared in the training subset, partition around medoids with cosine distance between observations [@Drost2018; @Schubert2019] demonstrated good explanatory power and superior reproducibility in cross-validation [@Lange2004] (__Supplementary Figure S2A__). 
Three clusters of participants named neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder) cluster after the patterns of their characteristic mental features were identified in the training subset. 
Subsequently, the mental cluster assignment could be robustly validated in the test subset as evident from comparable fractions of explained clustering variance in both study cohort subsets (training: `r signif(semi_clust$variance$variance[1], 2)`, test: `r signif(semi_clust$variance$variance[2], 2)`) and comparable cluster sizes (__Supplementary Figures S1B__ - __S4__).

Roughly one-third study participants assigned to the neutral clusters were characterized by low scores of anxiety, depression, panic and somatic symptoms along with high rating of coherence, resilient coping and quality of life. 
Scores of post-traumatic growth and post-traumatic stresswere the lowest in the neutral cluster. 
The PTG cluster demonstrated similarly low score of major mental health problems as neutral cluster individuals. 
By contrast, the overall measures of post-traumatic growth and its domains were the highest in the PTG cluster. 
In addition, ratings of post-traumatic stress disorder symptoms were slightly elevated in the PTG as compared with the neutral cluster. 
The remaining participants subsumed in the PTSD cluster displayed the most intense signs of anxiety, depression, panic and somatic symptoms as well as poor sense of coherence, resilient coping and low quality of life as compared with the neutral and PTG cluster. 
PTSD cluster was hallmarked by the highest scores of post-traumatic stress disorder symptoms in every PCL-5 DSM-5 domain. 
Post-traumatic growth was also present in the PTSD cluster as evident from elevated PTGI scores (__Figure \@ref(fig:fig-clust-psych)__, __Supplementary Figure S4__, __Supplementary Table S6__).

Along this line, clinically relevant symptoms of anxiety, depression and somatization were present virtually only in the PTSD clusters of the training and test subsets of the study cohort. 
Furthermore, frequencies of low and moderate resilient coping classes were by far the highest in the PTSD cluster. 
Finally, approximately 40% of PTSD cluster individuals were screened positive for at least one post-traumatic stress disorder symptom which was more than in the PTG 
(training: `r signif(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)), 2)[2] * 100`%, 
test: `r signif(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'PTG') %>% .$dsm5_cluster_class)), 2)[2] * 100`%) or neutral cluster 
(training: `r signif(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$training %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)), 2)[2] * 100`%, 
test: `r signif(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)/sum(table(feat_clust$ident_tbl$test %>% filter(clust_id == 'neutral') %>% .$dsm5_cluster_class)), 2)[2] * 100`%). 
An analogical trend was observed for the remaining post-traumatic stress disorder domains (__Figure \@ref(fig:fig-clust-symptoms)__).

@@

In individuals surveyed on average 3.5 years after the non-professional sport accident, some of its consequences like somatic health problems, unwilling flashbacks as well as phenomena of post-traumatic growth and symptoms of post-traumatic stress disorder were still tangible. 
By semi-supervised clustering we could discern three distinct patterns of mental health in this collective with each present in approximately one-third of participants: (1) a neutral mental response reaction, (2) salutatory reaction to the accident characterized by post-traumatic growth and (3) mental health deterioration hallmarked by symptoms of anxiety, depression, somatization, resilience loss and signs of post-traumatic stress disorder. 
Notably, separation of the PTG and PTSD clusters was in part blurry. 
Symptoms of post-traumatic stresswere observed also in PTG cluster participants albeit less frequently than in the PTSD cluster. 
Conversely, mental health problems in PTSD cluster participants were paralleled by some signs of post-traumatic growth. 
In a detailed comparison of `r length(clust_bcg$variables)` factors between the clusters, no distinct demographic, socioeconomic, clinical or accident-related background of the mental clusters could be identified. 
We found a moderate enrichment of females, low-to-moderate income participants, non-academic graduates, chronic somatic and psychiatric conditions as well as individuals requiring hospitalization and surgery in the PTG and, in particular, in the PTSD cluster. 
Injury severity and localization or professional rescue rates indicative of potentially more severe accident were comparable between the clusters. 
Similarly, the cluster assignment was unlikely affected by psychological support after the accident. 
In turn, the PTSD and, to a lesser extent, the PTG cluster were characterized by somatic health consequences of the accident, flashbacks, more caution during sport activity and subjective need for psychological support. 
Finally, the lacking non-mental explanatory variables strongly associated with the cluster assignment likely resulted in sub-optimal performance of machine learning classifiers employed for prediction of the cluster assignment and identification of accident victims at risk of mental health problems.

@@

Our study bears limitations. 
First, although the key phenomena phenomena in the newly identified mental clusters, post-traumatic growth and post-traumatic stress disorder, are clearly associated with trauma [@Tedeschi1996; @Bovin2016], we could not exclude the mental cluster assignment to be affected by other events than the sport accident. 
To account for that, follow-up studies including a control group are needed. 
Second, our variable set misses potentially important explanatory factors for the mental cluster classification such as length of hospital stay, rehabilitation need or ability to work. 
Finally, effects of the accident, injury severity, hospitalization or surgery on mental health after an accident may have been obscured in the cross-sectional cohort and needs to be investigated in a more defined collective e.g. of high risk sport accident victims or hospitalized individuals.

@@

tables$cohort_mental %>% 
  map_dfc(stri_replace, fixed = '\nrange', replacement = ', range') %>% 
  flextable %>% 
  flextable::width(1, width = 5.5, unit = 'cm') %>% 
  flextable::width(2, width = 7.5, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations. Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 7-nearest neighbors classifier. Three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTSD (post-traumatic stress disorder), All clustering parameters differed significantly between the clusters as assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic. Mean normalized scores of major mental health readouts (A), diminished quality of life (B), post-traumatic growth (C) and post-traumatic symptom disorder (D) in the mental clusters of the training and test subsets of the study cohort are depicted as solid lines. Tinted regions represent 2 $\times$ SEM (standard error of the mean). Numbers of observations in the clusters are displayed in the plot captions. GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._

@@

_Distribution of gender (A), annual household income (B), and frequencies of self-reported pre-existing somatic conditions (C) and self-reported pre-existing mental illness (D) in the mental clusters. Statistical significance was determined by $\chi^2$ test with Cramer V effect size statistic. Percentages in the mental clusters in the training and test subset of the study cohort are presented in stack plots. Effect sizes and p-values are displayed in the plot captions. Numbers of observations in the clusters are presented in the X axes._
